\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-03-29 paper.qmd
Correct the reference not rendering in Background section

The alternative approach to visualizing the high-dimensional data is to use linear projections. PCA is the classical approach, resulting in a set of new variables which are linear combinations of the original variables. Tours, defined by @Asimov1985, broaden the scope by providing movies of linear projections, that provide views the data from all directions. @lee2021review provides an review of the main developments in tours. There are many tour algorithms implemented, with many available in the R package `tourr` [@article61], and versions enabling better interactivity in `langevitour` [@article09] and `detourr` [@article22]. Linear projections are a safe way to view high-dimensional data, because they do not warp the space, so they are more faithful representations of the structure. 
However, linear projections can be cluttered, and global patterns can obscure local structure. The simple activity of projecting data from $p-D$ suffers from piling [@laa2022], where data concentrates in the center of projections. NLDR is designed to escape these issues, to exaggerate structure so that it can be observed. But as a result NLDR can hallucinate wildly, to suggest patterns that are not actually present in the data.

The solution is to use the tour to examine how the NLDR is warping the space. This approach follows what @article08 describes as *model-in-the-data-space*. The fitted model should be overlaid on the data, to examine the fit relative the spread of the observations. While this is straightforward, and commonly done when data is 2D, it is also possible in $p-D$, for many models, when a tour is used.

@article08 provides several examples of models overlaid on the data in $p-D$. In hierarchical clustering, a representation of the dendrogrom using points and lines can be constructed by augmenting the data with points marking merging of clusters. Showing the movie of linear projections reveals shows how the algorithm sequentially fitted the cluster model to the data. For linear discriminant analysis or model-based clustering the model can be indicated by $(p-1)-D$ ellipses. It is possible to see whether the elliptical shapes appropriately matches the variance of the relevant clusters, and to compare and contrast different fits. For PCA, one can display the $k-D$ plane of the reduced dimension using wireframes of transformed cubes. Using a wireframe is the approach we take here, to represent the NLDR model in $p-D$.

>>>>

The alternative approach to visualizing the high-dimensional data is to use linear projections. PCA is the classical approach, resulting in a set of new variables which are linear combinations of the original variables. Tours, defined by @Asimov1985, broaden the scope by providing movies of linear projections, that provide views the data from all directions. @lee2021review provides an review of the main developments in tours. There are many tour algorithms implemented, with many available in the R package `tourr` [@wickham2011], and versions enabling better interactivity in `langevitour` [@harisson2024] and `detourr` [@hart2022]. Linear projections are a safe way to view high-dimensional data, because they do not warp the space, so they are more faithful representations of the structure. 
However, linear projections can be cluttered, and global patterns can obscure local structure. The simple activity of projecting data from $p-D$ suffers from piling [@laa2022], where data concentrates in the center of projections. NLDR is designed to escape these issues, to exaggerate structure so that it can be observed. But as a result NLDR can hallucinate wildly, to suggest patterns that are not actually present in the data.

The solution is to use the tour to examine how the NLDR is warping the space. This approach follows what @wickham2015 describes as *model-in-the-data-space*. The fitted model should be overlaid on the data, to examine the fit relative the spread of the observations. While this is straightforward, and commonly done when data is 2D, it is also possible in $p-D$, for many models, when a tour is used.

@wickham2015 provides several examples of models overlaid on the data in $p-D$. In hierarchical clustering, a representation of the dendrogrom using points and lines can be constructed by augmenting the data with points marking merging of clusters. Showing the movie of linear projections reveals shows how the algorithm sequentially fitted the cluster model to the data. For linear discriminant analysis or model-based clustering the model can be indicated by $(p-1)-D$ ellipses. It is possible to see whether the elliptical shapes appropriately matches the variance of the relevant clusters, and to compare and contrast different fits. For PCA, one can display the $k-D$ plane of the reduced dimension using wireframes of transformed cubes. Using a wireframe is the approach we take here, to represent the NLDR model in $p-D$.

### 2024-03-29 paper.tex
Correct the reference not rendering in Background section

The alternative approach to visualizing the high-dimensional data is to
use linear projections. PCA is the classical approach, resulting in a
set of new variables which are linear combinations of the original
variables. Tours, defined by \citet{Asimov1985}, broaden the scope by
providing movies of linear projections, that provide views the data from
all directions. \citet{lee2021review} provides an review of the main
developments in tours. There are many tour algorithms implemented, with
many available in the R package \texttt{tourr} \citep{article61}, and
versions enabling better interactivity in \texttt{langevitour}
\citep{article09} and \texttt{detourr} \citep{article22}. Linear
projections are a safe way to view high-dimensional data, because they
do not warp the space, so they are more faithful representations of the
structure. However, linear projections can be cluttered, and global
patterns can obscure local structure. The simple activity of projecting
data from \(p-D\) suffers from piling \citep{laa2022}, where data
concentrates in the center of projections. NLDR is designed to escape
these issues, to exaggerate structure so that it can be observed. But as
a result NLDR can hallucinate wildly, to suggest patterns that are not
actually present in the data.

The solution is to use the tour to examine how the NLDR is warping the
space. This approach follows what \citet{article08} describes as
\emph{model-in-the-data-space}. The fitted model should be overlaid on
the data, to examine the fit relative the spread of the observations.
While this is straightforward, and commonly done when data is 2D, it is
also possible in \(p-D\), for many models, when a tour is used.

\citet{article08} provides several examples of models overlaid on the
data in \(p-D\). In hierarchical clustering, a representation of the
dendrogrom using points and lines can be constructed by augmenting the
data with points marking merging of clusters. Showing the movie of
linear projections reveals shows how the algorithm sequentially fitted
the cluster model to the data. For linear discriminant analysis or
model-based clustering the model can be indicated by \((p-1)-D\)
ellipses. It is possible to see whether the elliptical shapes
appropriately matches the variance of the relevant clusters, and to
compare and contrast different fits. For PCA, one can display the
\(k-D\) plane of the reduced dimension using wireframes of transformed
cubes. Using a wireframe is the approach we take here, to represent the
NLDR model in \(p-D\).

>>>>

The alternative approach to visualizing the high-dimensional data is to
use linear projections. PCA is the classical approach, resulting in a
set of new variables which are linear combinations of the original
variables. Tours, defined by \citet{Asimov1985}, broaden the scope by
providing movies of linear projections, that provide views the data from
all directions. \citet{lee2021review} provides an review of the main
developments in tours. There are many tour algorithms implemented, with
many available in the R package \texttt{tourr} \citep{wickham2011}, and
versions enabling better interactivity in \texttt{langevitour}
\citep{harisson2024} and \texttt{detourr} \citep{hart2022}. Linear
projections are a safe way to view high-dimensional data, because they
do not warp the space, so they are more faithful representations of the
structure. However, linear projections can be cluttered, and global
patterns can obscure local structure. The simple activity of projecting
data from \(p-D\) suffers from piling \citep{laa2022}, where data
concentrates in the center of projections. NLDR is designed to escape
these issues, to exaggerate structure so that it can be observed. But as
a result NLDR can hallucinate wildly, to suggest patterns that are not
actually present in the data.

The solution is to use the tour to examine how the NLDR is warping the
space. This approach follows what \citet{wickham2015} describes as
\emph{model-in-the-data-space}. The fitted model should be overlaid on
the data, to examine the fit relative the spread of the observations.
While this is straightforward, and commonly done when data is 2D, it is
also possible in \(p-D\), for many models, when a tour is used.

\citet{wickham2015} provides several examples of models overlaid on the
data in \(p-D\). In hierarchical clustering, a representation of the
dendrogrom using points and lines can be constructed by augmenting the
data with points marking merging of clusters. Showing the movie of
linear projections reveals shows how the algorithm sequentially fitted
the cluster model to the data. For linear discriminant analysis or
model-based clustering the model can be indicated by \((p-1)-D\)
ellipses. It is possible to see whether the elliptical shapes
appropriately matches the variance of the relevant clusters, and to
compare and contrast different fits. For PCA, one can display the
\(k-D\) plane of the reduced dimension using wireframes of transformed
cubes. Using a wireframe is the approach we take here, to represent the
NLDR model in \(p-D\).
